Bootstrapping Multi-atlas Hippocampal Segmentation with \mb

### 2013-05-17 paper.Rnw
Rewrote bits of the methods (MB description, etc.) for clarity and completeness.
Also added in a bland-altman plot for ADNI1-1yr dataset

%notes and TODO formatting
%\newcommand{\marginnote}[1]{\-\marginpar[\raggedleft\footnotesize #1]{\raggedright\footnotesize #1}}
%\newcommand{\comment}[1]{\begin{kframe}{\textcolor{red}{#1}}\end{kframe}}
%\newcommand{\todo}[1]{\comment{TODO #1}}
%\newcommand{\mc}[1]{\comment{MC: #1}}
\begin{document}

\title{Bootstrapping Multi-atlas Hippocampal Segmentation with \mb}
\author{Pipitone J., Winterburn J., Lerch J., Pruessner J., Lepage M., \\ 
Voineskos A., Chakravarty M.M., and \\ 
the Alzheimer's Disease Neuroimaging Initiative}
\maketitle

@@

\subsection{The \mb Algorithm}
In this paper we will make a distinction between an atlas and a template image.
The term {\it atlas} refers to two image volumes: an intensity image ({\it atlas
image}) and a corresponding manual segmentation image ({\it atlas labels}). The
term {\it template} refers to the {\it use} an any image and corresponding label
image (whether the labels are manual or computed) as a model in the segmentation
of another image.  The terms {\it atlas library} and {\it template library} mean
a set of such images.  Additionally, we will use the terms {\it target} to refer
to an intensity image for which we would like an segmentation.

The simplest form of multi-atlas segmentation combines labellings derived from
several atlases by way of label fusion \citep{Heckemann2006,Heckemann2011}. We
will refer to this as {\em basic multi-atlas segmentation}. The schema as for
this method is as follows: 
\begin{enumerate}
\item   An atlas library and set of target images are given as input. The atlas
        library is used as a template library in the following steps; 
\item   Each atlas intensity image is nonlinearly registered to each target 
        intensity image; 
\item   Label images from each atlas are propagated via the resulting
        transformations to the target image space; and 
\item   the resulting labels are fused to produce a single, definitive
        segmentation. 
\end{enumerate}       
The particular registration and voting method used are left unspecified.

\mb is best understood as an extension of the basic multi-atlas
segmentation schema. Instead of using the atlas library to directly label the
target images, a subset of the input images are selected as template images and
then labelled.  The choice of targets used in the template library can be made
to reflect the neuroanatomy or demographics of the target set as a whole (for
instance, by sampling equally from cases and controls). Once the template
library images have been chosen, a truncated version of basic multi-atlas
segmentation is used to label the template library images without performing
label fusion. Instead, each template image receives multiple labellings: one
from each atlas image. A second round of basic multi-atlas segmentation uses the
template library to segment the entire set of target images (including those
images used in the template library). Label fusion in this final step fuses all
labels from all templates. To summarize, figure \ref{alg:MAGeT} describes the
\mb algorithm in pseudocode.

Source code for \mb can be found at
\url{http://github.com/pipitone/MAGeTbrain}.

@@

\subparagraph{Label fusion methods}
Label fusion is a term given to the process of combining the information from
several candidate labellings for an intensity image into a single labelling.  In
this experiment we explore three fusion methods.

\begin{description}
  \item[Voxel-wise Majority Vote] \hfill \\
  Labels are propagated from all template library images to a target.  Each
  output voxel is given the most frequent label at that voxel location amongst
  all candidate labellings.  Ties are broken arbitrarily.

  \item[Cross-correlation Weighted Majority Vote] \hfill \\
  An optimal combination of targets from the template library has previously been
  shown to improve segmentation accuracy \citep{Aljabar2009,Collins2010}.  In this
  method, each template library image is ranked in similarity to each unlabelled 
  image by the normalized cross-correlation (CC) of image intensities after linear
  registration, over a region of interest (ROI) generously encompassing the 
  hippocampus.  Only the top ranked template library image labels are used in a
  voxel-wise majority vote. The ROI is heuristically defined as the extent of all
  atlas labels after linear registration to the template, dilated by three voxels
  \citep{Chakravarty2012}.  The number of top ranked template library image labels
  is a configurable parameter.

@@

  \item[Normalised Mutual Information Weighted Majority Vote] \hfill \\
  This method is similar to cross-correlation weighted voting except that image
  similarity is calculated by the normalised mutual information score over the
  region of interest\citep{Studholme2001}.  The {\tt itk\_similarity} utility
  from the EZMinc toolkit\footnote{\url{https://github.com/vfonov/EZminc}} is
  used to calculate the normalised mutual information measure between to images.
\end{description}

@@

Reported subject segmentation accuracy is averaged over the ten validation
rounds for each parameter setting.

@@

In this experiment, the five $300\mu$-isotropic voxel Winterburn atlases are
used as the atlas library for \mb segmentation. LOOCV is conducted separately
for two different input datasets: the four Winterburn atlas subject BRAVO MR
images (referred to as the {\em WA BRAVO} dataset), and the five
Winterburn atlas subject images resampled to 0.9mm-isotropic voxel resolution
(referred to as the {\em WA Resampled} dataset. Each subject in the
dataset is segmented by \mb with that subject's image excluded from the atlas
library. The template library consists of 3T T1 images (0.9mm-isotropic
voxels) of healthy subjects in addition to the images from the dataset being
evaluated.

@@

\paragraph{Evaluation method}  
The segmentation volumes of each hippocampal subregion are compared to the
expert manual segmentations of the unmodified Winterburn atlases.  As a point of
comparison, we also calculate the subregion volumes from the Winterburn atlas
segmentations after downsampling to 0.9mm-isotropic voxels.

@@

\paragraph{Evaluation method}
The manual segmentation protocol used to segment the Winterburn atlases is
similar to, but different from, the protocol used to segment the SZFEP dataset.
Therefore, rather than use an overlap metric, \mb hippocampal volumes are
compared to the corresponding manual segmentation volumes.

\todo{discuss Bland-Altman?  Or bias analysis}

@@

Finally, \mb segmentation volumes show differential agreement with manual (SNT) 
volumes throughout the range of hippocampal volumes (Figure 
\ref{ADNI1-xval-Bland-Altman}). \mb shows a negligble fixed bias towards 
producing smaller segmentations, and shows a slight proportional bias towards 
smaller segmentations with larger hippocampi.
\todo{Note limits of agreement?}

\todo{Include multi-atlas 9-atlas in plot as a comparison}

@@

stats.tests = ddply(data.stats, c("atlases","templates.mb"), function (df) {
  t.var = t.test(df$MA.var, df$MB.var)
  t.sd  = t.test(df$MA.sd , df$MB.sd)
  data.frame(
    stat      = c("Variance", "Standard Deviation"),
    statistic = c(t.var$statistic, t.sd$statistic),
    p.value   = c(t.var$p.value, t.sd$p.value),
    direction = c(t.var$statistic < 0, t.sd$statistic < 0)
  )
})

ggplot(stats.tests, 
  aes(x=templates.mb, y=p.value, colour=as.factor(atlases), 
      size = abs(statistic))) + 
  facet_grid( . ~ stat) + 
  geom_line(aes(size=0.1,area=0.1)) + 
  geom_point(aes(shape = factor(direction))) + 
  geom_hline(aes(yintercept=0.05, alpha=0.5), linetype='dashed') + 
  geom_hline(aes(yintercept=0.01, alpha=0.5), linetype='dashed') + 
  scale_shape_discrete("t-statistic sign", labels=c("positive","negative")) + 
  scale_size_area(name="t-statistic magnitude") + 
  scale_x_continuous(breaks=seq(1,20,by=2)) + 
  scale_y_log10(limits=c(0.001,1), breaks=c(0.01,0.05,0.1,1.0)) + 
  xlab( "Number of Templates" ) + 
  ylab( "p-value" ) +
  scale_colour_hue(name="Atlases") +
  scale_alpha_continuous(guide='none') + 
  scale_linetype_discrete(guide='none') + 
  scale_size_continuous(guide='none') + 
  opts(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{{\bf Difference in variability of \mb vs. multi-atlas segmentation 
  accuracy.}
  Variability of segmentation accuracy within 10 rounds of validation per subject. 
  Standard deviation and variance is computed per subject for both \mb and 
  multi-atlas, and the distribution of these test statistics is compared with a 
  t-test. The p-value of this test, is shown on the y-axis (scaled logarithmically).}
  \label{ADNI1-xval-variability}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: Bland-Altman plots
\begin{figure}
<<ADN1-xval-Bland-Atlman, cache=TRUE, dependson='ADNI1-Xval-prep', fig.width=8, fig.height=8>>=
df_subset = subset(all_data, 
                      reg_method   == "ANTS" &
                      method.mb    == "Majority Vote" &
                      templates.mb %in% c(1,11,19) &
                      atlases      %in% c(1,5,9) &
                      volume.ma     < 10000)

# get SNT data formatted like 
snt   = read.csv("data/a2a_snt_volumes.csv")
df_merged = merge(df_subset, snt, all.x = TRUE)

hc_vol = ddply(df_merged, c("subject", "label", "atlases", "templates.mb"), function (df) {
  data.frame(
    snt = sum(df$volume.snt),
    ma = sum(df$volume.ma),
    mb = sum(df$volume.mb)
  )
})
hc_vol$diff = hc_vol$snt - hc_vol$mb

group.stats = ddply(hc_vol, c("atlases", "templates.mb"), function (df) {
  data.frame(
    mean = mean(df$diff),
    sd   = sd(df$diff)
  )
})
hc_vol = merge(hc_vol, group.stats)
hc_vol$Atlases = hc_vol$atlases
hc_vol$Templates = hc_vol$templates.mb

ggplot(hc_vol, aes(x=ma, y = diff, group=Atlases)) + 
    geom_hline(aes(yintercept = mean + c(-2, 0, 2) * sd, colour='brown'), 
                linetype=2) + 
    geom_point() + 
    geom_smooth(method="lm") + 
  facet_grid(Atlases~Templates, labeller = label_both) + 
  xlab(expression('Mean manual and MAGeT-Brain total hippocampus volume ' (mm^3))) + 
  ylab(expression('Manual - MAGeT-Brain total hippocampus volume ' (mm^3))) 
@
  \caption{Bland-Altman plots comparing manual and \mb hippocampal volumes 
  when using a varying number of atlases and templates.}
  \label{ADNI1-xval-Bland-Altman}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

snt.tests <- ddply(totals.complete, c("DX"), function(df) {
  round(data.frame( MAGeT = cor(df$SNT, df$MAGeT),
              MAPER = cor(df$SNT, df$MAGeT),
              FS    = cor(df$SNT, df$FS),
              FSL   = cor(df$SNT, df$FSL)
              ), 2)
})

latex(snt.tests, file="", 
      size="scriptsize",
      caption="", row.names="",
      title="")
@

>>>>

%notes and TODO formatting
%\newcommand{\marginnote}[1]{\-\marginpar[\raggedleft\footnotesize #1]{\raggedright\footnotesize #1}}
%\newcommand{\comment}[1]{\begin{kframe}{\textcolor{red}{#1}}\end{kframe}}
%\newcommand{\todo}[1]{\comment{TODO #1}}
%\newcommand{\mc}[1]{\comment{MC: #1}}
\begin{document}
%\SweaveOpts{concordance=TRUE}

\title{Bootstrapping Multi-atlas Hippocampal Segmentation with \mb}
\author{Pipitone J., Winterburn J., Lett, T., Lerch J., Pruessner J., Lepage M., \\ 
Voineskos A., Chakravarty M.M., and \\ 
the Alzheimer's Disease Neuroimaging Initiative}
\maketitle

@@

\subsection{The \mb Algorithm}
In this paper, we will use the term {\em label} to mean any segmentation (manual or 
derived) of an MR image.  {\em Label propagation}, or {\em labelling}, is the 
process by which two images are registered and the resulting transformation 
is applied to the labels from one image to bring them into alignment with the 
other image. We will use the term {\em atlas} to mean a manually segmented image,
and the term {\em template} to mean an automatically segmented image (i.e. 
via label propagation). The terms {\em atlas library} and {\em template library} 
describe any set of such images.Additionally, we will use the term {\em target} 
to refer to an unlabelled image that is undergoing segmentation.

The simplest form of multi-atlas segmentation, (which we will call {\em basic 
multi-atlas segmentation}), involves three steps.  First, each labelled 
input image (i.e. atlas or template) is registered to an unlabelled target image.
Second, the labels from each image are propogated to the target image space. 
Third, the labels are combined into a single labelling by way of a label fusion method 
\citep{Heckemann2006, Heckemann2011}. This method is described in detail in other
publications \citep{Collins2010,Heckemann2011,Aljabar2009}.

\mb bootstraps the creation of a large template library given a limited input atlas 
library, and then uses the template library in basic multi-atlas segmentation. 
Images for the template library are selected from a set of input target images, 
either arbitrarily or so as to reflect the neuroanatomy or demographics of the 
target set as a whole (for instance, by sampling equally from cases and controls). 
The template library images are then labelled by each of the atlases. Basic 
multi-atlas segmentation is then conducted using the template library
to segment the entire set of target images (including the targets whose images
are used in the construction of the template library).  Since each template
library image has multiple labels (one from each atlas), the final number of 
labels to be fused for each target may be quite large (i.e. \# of atlas $\times$ 
\# of templates).

Figure \ref{alg:MAGeT} describes the \mb algorithm in pseudocode. Source code 
for \mb can be found at \url{http://github.com/pipitone/MAGeTbrain}.

@@

\paragraph{Label fusion methods}
Label fusion is a term given to the process of combining the information from
several candidate labellings for an intensity image into a single labelling.  In
this experiment we explore three fusion methods.

\subparagraph{Voxel-wise Majority Vote}
  Labels are propagated from all template library images to a target.  Each
  output voxel is given the most frequent label at that voxel location amongst
  all candidate labellings.  Ties are broken arbitrarily.

\subparagraph{Cross-correlation Weighted Majority Vote}
  An optimal combination of targets from the template library has previously been
  shown to improve segmentation accuracy \citep{Aljabar2009,Collins2010}.  In this
  method, each template library image is ranked in similarity to each unlabelled 
  image by the normalized cross-correlation (CC) of image intensities after linear
  registration, over a region of interest (ROI) generously encompassing the 
  hippocampus.  Only the top ranked template library image labels are used in a
  voxel-wise majority vote. The ROI is heuristically defined as the extent of all
  atlas labels after linear registration to the template, dilated by three voxels
  \citep{Chakravarty2012}.  The number of top ranked template library image labels
  is a configurable parameter.

@@

\subparagraph{Normalised Mutual Information Weighted Majority Vote}
  This method is similar to cross-correlation weighted voting except that image
  similarity is calculated by the normalised mutual information score over the
  region of interest\citep{Studholme2001}.  The {\tt itk\_similarity} utility
  from the EZMinc toolkit\footnote{\url{https://github.com/vfonov/EZminc}} is
  used to calculate the normalised mutual information measure between to images.

@@

The manual segmentations (SNT) provided as part of the ADNI dataset are used as 
the gold standard to compare with. We compare both \mb labels and multi-atlas 
labels.  The segmentation accuracy reported is averaged over the ten validation
rounds for each parameter setting.

In order to investigate the performance of \mb in a real world setting in which 
only one set of atlas and template images are used, we explore the variability 
in label agreement at fixed parameter settings when the choice for atlas and 
template images is varied. This is achieved by first computing the standard 
deviation and variance of DSC scores in each block 
of ten validation rounds per subject. The distribution of these statistics across 
all subjects is then compared between \mb and multi-atlas using a Student's 
t-test. A significant difference between distributions is taken to show either
a larger or smaller level of variability between methods.

@@

In this experiment, the five $300\mu$-isotropic voxel Winterburn atlases are
used as the atlas library for \mb segmentation. LOOCV is conducted separately
for two different input datasets: the four Winterburn atlas subject T1 BRAVO
images (referred to as the {\em BRAVO} dataset), and the five
Winterburn atlas subject images resampled to 0.9mm-isotropic voxel resolution
(referred to as the {\em Resampled} dataset. Each subject in the
dataset is segmented by \mb with that subject's image excluded from the atlas
library. The template library consists of 3T T1 images (0.9mm-isotropic
voxels) of healthy subjects in addition to the images from the dataset being
evaluated.

@@

\paragraph{Evaluation method}  
\mb performance is measured with respect to straight-forward nearest-neighbour
resampling of the Winterburn atlas labels to 0.9mm-isotropic voxels (i.e. no 
label propagation). The volumes of each hippocampal subregion in the downsampled
Winterburn atlas labels are compared to the volumes obtained from the full 
resolution Winterburn atlas labels, the \mb labels on the downsampled Winterburn 
atlas images, and the \mb labels on the T1 BRAVO acquisitions.

@@

\paragraph{Evaluation method}
The manual segmentation protocol used to segment the Winterburn atlases is
similar to, but different from, the protocol used to segment the SZFEP dataset.
Therefore, rather than evaluate using an overlap metric, such as DSC, which 
assumes a fixed segmentation protocol, we instead correlate whole hippocampus 
volume between \mb and manual segmentations.

@@

@@

stats.tests = ddply(data.stats, c("atlases","templates.mb"), function (df) {
  t.var = t.test(df$MA.var, df$MB.var)
  t.sd  = t.test(df$MA.sd , df$MB.sd)
  diff_mean_var = mean(df$MA.var - df$MB.var)
  diff_mean_sd  = mean(df$MA.sd -  df$MB.sd)

  # test statistic is positive if MA mean is larger than MB
  # i.e. that MA has a greater mean of variances/SDs
  # i.e. that MB has a smaller mean, and so is doing "better"
  data.frame(
    stat      = c("Variance", "Standard Deviation"),
    statistic = c(t.var$statistic, t.sd$statistic),
    p.value   = c(t.var$p.value, t.sd$p.value),
    direction = c(diff_mean_var, diff_mean_sd)
  )
})

ggplot(subset(stats.tests, direction > 0), 
  aes(x=templates.mb, y=p.value, colour=as.factor(atlases))) + 
  facet_grid( . ~ stat) + 
  geom_line(size=1) +
  geom_point(size=3) + 
  geom_hline(aes(yintercept=0.05, alpha=0.5), linetype='dashed') + 
  geom_hline(aes(yintercept=0.01, alpha=0.5), linetype='dashed') + 
  scale_x_continuous(breaks=seq(1,20,by=2)) + 
  scale_y_log10(limits=c(0.001,1), breaks=c(0.01,0.05,0.1,1.0)) + 
  xlab( "Number of Templates" ) + 
  ylab( "p-value" ) +
  scale_colour_hue(name="Atlases") +
  scale_alpha_continuous(guide='none') + 
  scale_linetype_discrete(guide='none') + 
  scale_size_continuous(guide='none') + 
  theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{{\bf Difference in variability of \mb vs. multi-atlas segmentation 
  accuracy.}
  Variability of segmentation accuracy within 10 rounds of validation per subject. 
  Standard deviation and variance is computed per subject for both \mb and 
  multi-atlas, and the distribution of these test statistics is compared with a 
  t-test. The p-value of this test, is shown on the y-axis (scaled logarithmically).
  Only points where \mb mean variability is lower multi-atlas are shown.}
  \label{ADNI1-xval-variability}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Additionally \mb volumes show a fixed bias towards larger volumes by an 
average $-205.2mm^3$, and a proportional bias towards producing smaller 
segmentations for larger hippocampi (Figure \ref{fig:ADNI1-src-Bland-Altman}.

@@

snt.tests <- ddply(totals.complete, c("DX"), function(df) {
  round(data.frame( MAGeT = cor(df$SNT, df$MAGeT),
              MAPER = cor(df$SNT, df$MAGeT),
              FS    = cor(df$SNT, df$FS),
              FSL   = cor(df$SNT, df$FSL)
              ), 2)
})
caption = paste(
  "Pearson correlation between manual (SNT) whole hippocampus volume", 
  "and whole hippocampus volume by method.", sep=" ")
latex(snt.tests, file="", 
      size="scriptsize",
      caption=caption, row.names="",
      title="")
@

\begin{figure}
<<ADN1-src-Bland-Atlman, cache=TRUE, dependson='ADNI-baseline-volumes-prep', fig.width=4, fig.height=4>>=
melted=melt(totals.complete, measure.vars=c("FS", "FSL", "MAPER","MAGeT"), 
            variable.name = "method", value.name = "volume")
melted=subset(melted, method=="MAGeT")

bland_altman_plot <- function(x, y) {
  data = data.frame(x=x,y=y,diff=x-y)
  upper = mean(data$diff) + 2 * sd(data$diff)
  lower = mean(data$diff) - 2 * sd(data$diff)
  return(ggplot(data, aes(x= x, y = diff)) + 
    geom_hline(yintercept = mean(data$diff) + c(-2, 0, 2) * sd(data$diff), 
               linetype=2, color='brown') + 
    annotate("text", y = upper, x=Inf, 
              label=paste("std =",round(upper,0)), 
              hjust=1.1,vjust=1.5) + 
    annotate("text", y = lower, x=Inf, 
              label=paste("-std =",round(lower,0)), 
              hjust=1.1,vjust=1.5) + 
    geom_point())
}

melted$diff = melted$SNT - melted$volume
limits = mean(melted$diff) + c(-2, 0, 2) * sd(melted$diff)

ggplot(melted, aes(x= SNT, y = diff)) + 
  geom_hline(yintercept = limits, linetype=2, color='brown') + 
 annotate("text", y = limits, x=Inf, label=round(limits,1), 
            hjust=1.1,vjust=-0.8, size=3) + 
  geom_point() + 
  xlab('Mean manual and MAGeT Brain total Hippocampus volume') + 
  ylab('Manual - MAGeT Brain volume') + 
  geom_smooth(method="lm", formula = y~x)
@
  \caption{Bland-Altman plot comparing manual (SNT) and \mb hippocampal volumes 
  in the ADNI1 1.5T 1Yr Complete dataset.}
  \label{fig:ADNI-scr-Bland-Altman}
\end{figure}
